---
title: "Complete LANTERN workflow: split → GDS → test → combine"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Complete LANTERN workflow: split → GDS → test → combine}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(lantern)
library(SeqArray)
```

## Overview

This vignette walks through the complete LANTERN pipeline from raw phased
VCF + RFMix MSP files to ancestry-stratified gene-level burden test p-values.

```
Phased VCF (BCF) + RFMix MSP
          │
          ▼
  run_combined_pipeline()
          │
   ┌──────┴──────┐
phased split  unphased split
   │                 │
   ▼                 ▼
write_dosage_gds()  write_dosage_gds()
   │                 │
 AFR.gds  EUR.gds  AFR_un.gds  EUR_un.gds
          │
    glmmkin()  ← phenotype + kinship matrix
          │
    ┌─────┼─────┐
    │     │     │
 SMMAT  SMMAT  SMMAT
  AFR   EUR   observed
    │     │
  cauchy_combine()
          │
  per-gene multi-ancestry p-value
```

Steps 1–3 run on synthetic data and are fully executable.
Steps 4–5 (GMMAT/SMMAT) require the **GMMAT** package and are shown as
non-evaluated code blocks with expected output.

---

## 1  Synthetic dataset

We create a minimal phased VCF and RFMix MSP file in a temporary directory.
In a real analysis replace these paths with your cohort files.

```{r setup-data}
set.seed(42)
tmp <- tempfile(); dir.create(tmp)
vcf_path <- file.path(tmp, "cohort_phased.vcf")
msp_path <- file.path(tmp, "cohort_rfmix.msp.tsv")

# 8 variants × 5 phased samples
samples <- paste0("S", 1:5)
header <- c(
  "##fileformat=VCFv4.1",
  '##FORMAT=<ID=GT,Number=1,Type=String,Description="Genotype">',
  paste(c("#CHROM","POS","ID","REF","ALT","QUAL","FILTER","INFO","FORMAT",
          samples), collapse = "\t")
)
variants <- rbind(
  c("19","1100",".","A","G","100","PASS",".","GT","0|1","1|0","0|0","1|0","0|1"),
  c("19","1200",".","C","T","100","PASS",".","GT","1|1","0|1","1|0","0|0","1|1"),
  c("19","1700",".","G","A","100","PASS",".","GT","0|0","1|0","0|1","1|0","0|0"),
  c("19","2100",".","T","C","100","PASS",".","GT","1|0","0|1","1|1","0|1","1|0"),
  c("19","2300",".","A","C","100","PASS",".","GT","0|0","0|0","0|1","0|0","0|0"),
  c("19","2800",".","C","G","100","PASS",".","GT","1|1","0|0","1|0","0|1","1|0"),
  c("19","3100",".","G","T","100","PASS",".","GT","0|1","1|0","0|0","1|1","0|1"),
  c("19","3400",".","T","A","100","PASS",".","GT","1|0","0|1","0|0","0|0","1|1")
)
writeLines(c(header, apply(variants, 1, paste, collapse = "\t")), vcf_path)

# RFMix MSP: two ancestry tracts (AFR=0, EUR=1)
#   tract 1 [1000,2500): S1=AFR|EUR  S2=EUR|AFR  S3=AFR|EUR  S4=EUR|EUR  S5=AFR|AFR
#   tract 2 [2500,4000): S1=EUR|AFR  S2=AFR|EUR  S3=EUR|AFR  S4=AFR|EUR  S5=EUR|EUR
msp_lines <- c(
  paste(c("#Subpopulation order/codes: AFR=0", "EUR=1"), collapse = "\t"),
  paste(c("#chm","spos","epos","sgpos","egpos","n snps",
          "S1.0","S1.1","S2.0","S2.1","S3.0","S3.1",
          "S4.0","S4.1","S5.0","S5.1"), collapse = "\t"),
  paste(c("19","1000","2500","0.00","0.05","5",
          "0","1","1","0","0","1","1","1","0","0"), collapse = "\t"),
  paste(c("19","2500","4000","0.05","0.12","3",
          "1","0","0","1","1","0","0","1","1","1"), collapse = "\t")
)
writeLines(msp_lines, msp_path)

cat("VCF:  ", vcf_path, "\n")
cat("MSP:  ", msp_path, "\n")
```

---

## 2  Step 1: Split genotypes by ancestry

`run_combined_pipeline()` parses the VCF and MSP **once** and returns both
phased (deterministic) and unphased (proportional) dosage matrices for every
population found in the MSP.

```{r split, message=FALSE}
combined <- run_combined_pipeline(
  vcf_path = vcf_path,
  msp_path = msp_path,
  verbose  = FALSE
)

# Population names are taken directly from the MSP header
pop_names <- setdiff(names(combined$phased),
                     c("variant_info", "sample_ids", "overlap"))
cat("Populations:", paste(pop_names, collapse = ", "), "\n")
cat("Samples:    ", length(combined$phased$sample_ids), "\n")
cat("Variants (phased split):", nrow(combined$phased$variant_info), "\n")
```

```{r inspect-split}
cat("--- Phased AFR dosage (variants × samples) ---\n")
print(combined$phased$AFR)

cat("\n--- Phased EUR dosage ---\n")
print(combined$phased$EUR)

cat("\n--- Unphased AFR dosage ---\n")
print(combined$unphased$AFR)

# Sanity: phased and unphased totals should be equal
total_ph <- Reduce("+", combined$phased[pop_names])
total_un <- Reduce("+", combined$unphased[pop_names])
cat("\nMax absolute difference between phased and unphased totals:",
    max(abs(total_ph - total_un)), "\n")
```

The key difference between the two splits:

| | Phased | Unphased |
|---|---|---|
| **Input** | per-haplotype GT + ancestry | diploid GT + diploid ancestry code |
| **Method** | `split_phased_multi()` | `split_by_ancestry_multi()` |
| **Accuracy** | Exact (deterministic) | Proportional (singletons → p=0.5 each) |
| **Typical use** | Fine-mapping, haplotype tests | Standard burden tests |

---

## 3  Step 2: Write ancestry-specific GDS files

`write_dosage_gds()` wraps the VCF → GDS conversion.  Each population gets its
own GDS file which will be passed to SMMAT.

```{r write-gds, message=FALSE}
gds_dir <- file.path(tmp, "gds")
dir.create(gds_dir)

# Write one GDS per population for the phased split
vi   <- combined$phased$variant_info   # chrom / pos / ref / alt
ids  <- combined$phased$sample_ids

gds_paths <- list()
for (pop in pop_names) {
  gds_paths[[pop]] <- file.path(gds_dir, paste0(pop, "_phased.gds"))
  write_dosage_gds(combined$phased[[pop]], vi, ids, gds_paths[[pop]])
}

cat("GDS files written:\n")
for (pop in pop_names) cat(" ", pop, "->", gds_paths[[pop]], "\n")
```

Verify one file contains the expected variant and sample counts:

```{r verify-gds}
gds <- seqOpen(gds_paths[["AFR"]])
cat("Variants in AFR GDS:", seqSummary(gds, check = "none")$num.variant, "\n")
cat("Samples  in AFR GDS:", seqSummary(gds, check = "none")$num.sample,  "\n")
seqClose(gds)
```

---

## 4  Step 3: Gene-group file

SMMAT reads variant membership from a tab-delimited file (no header) with
columns: **gene, chrom, pos, ref, alt, weight**.

```{r gene-group}
# Assign all 8 variants to one toy gene
vi$gene   <- "MYGENE1"
vi$weight <- 1   # equal weights; use Beta(1,25) MAF weights for SKAT-O

gene_group_file <- file.path(tmp, "gene_groups.tsv")
write.table(vi[, c("gene","chrom","pos","ref","alt","weight")],
            gene_group_file, sep = "\t",
            row.names = FALSE, col.names = FALSE, quote = FALSE)

cat("Gene group file (all rows):\n")
print(read.table(gene_group_file, sep = "\t",
                 col.names = c("gene","chrom","pos","ref","alt","weight")))
```

---

## 5  Step 4: Null model with GMMAT  *(requires GMMAT)*

The null mixed model is fit **once** and shared across all SMMAT calls for
every population and every gene.

```{r null-model, eval=FALSE}
library(GMMAT)

# In real data replace diag(N) with your kinship / GRM matrix
N      <- length(ids)
Kmat   <- diag(N)
rownames(Kmat) <- colnames(Kmat) <- ids

# Simulate a case/control phenotype
set.seed(1)
df_pheno <- data.frame(id = ids, y = rbinom(N, 1, 0.4))

model0 <- glmmkin(
  y ~ 1,
  data   = df_pheno,
  kins   = Kmat,
  id     = "id",
  family = binomial(link = "logit")
)
# For a quantitative trait: family = gaussian(link = "identity")
```

> **Tip**: On large cohorts, fitting `glmmkin` with a dense kinship matrix
> can take minutes.  Use the sparse `kins` argument (a `Matrix::sparseMatrix`)
> for faster computation.

---

## 6  Step 5: SMMAT gene-level tests  *(requires GMMAT)*

Run SMMAT separately for each ancestry and for the raw observed genotype.

```{r smmat, eval=FALSE}
run_smmat <- function(gds_file) {
  SMMAT(
    obj.nullmodel = model0,
    gds.fn        = gds_file,
    group.file    = gene_group_file,
    MAF.range     = c(0, 0.05),   # rare variants only
    miss.cutoff   = 1,            # keep variants with any missing rate
    method        = "davies",
    is.dosage     = TRUE,         # required for ancestry-specific dosage GDS
    ncores        = 1L
  )
}

# Run for each ancestry
smmat_results <- lapply(gds_paths, run_smmat)

# For comparison: observed (raw unphased) dosage — requires a separate GDS
# built with the full diploid genotype matrix
# smmat_obs <- SMMAT(..., is.dosage = FALSE)
```

Each result is a data frame with one row per gene and columns
`gene`, `N.variant`, `N.sample`, `B`, `B.var`, `Q_B`, `p_B`, `Q_S`, `p_S`,
`p_SMMAT` (omnibus p-value).

Example output:
```
     gene N.variant N.sample          B      B.var       Q_B       p_B
1 MYGENE1         8        5  0.0312148 0.00128942  7.554824 0.0059940
       Q_S        p_S    p_SMMAT
1 7.554824 0.00599399 0.00599399
```

---

## 7  Step 6: Cauchy combination

`cauchy_combine()` combines the per-ancestry SMMAT p-values for each gene into
a single multi-ancestry p-value.  The test remains valid even when the
ancestries are correlated (as they always are in admixed cohorts).

```{r cauchy-demo}
# Toy example: two-population combination
p_afr <- 0.008   # hypothetical SMMAT p-value for AFR ancestry
p_eur <- 0.042   # hypothetical SMMAT p-value for EUR ancestry

p_multi <- cauchy_combine(c(p_afr, p_eur))
cat(sprintf("AFR p = %.4f  |  EUR p = %.4f  |  combined p = %.4f\n",
            p_afr, p_eur, p_multi))
```

```{r cauchy-table}
# Simulate results for five genes
set.seed(7)
n_genes <- 5
res_sim <- data.frame(
  gene   = paste0("GENE", seq_len(n_genes)),
  p_afr  = runif(n_genes, 1e-5, 0.5),
  p_eur  = runif(n_genes, 1e-5, 0.5)
)
res_sim$p_cauchy <- mapply(function(a, e) cauchy_combine(c(a, e)),
                           res_sim$p_afr, res_sim$p_eur)
print(res_sim)
```

```{r cauchy-3pop}
# Three-population example (AFR, EUR, NAT)
cauchy_combine(c(0.005, 0.04, 0.30))

# Weighted: give double weight to AFR if N_AFR > N_EUR
cauchy_combine(c(0.005, 0.04), weights = c(2, 1))
```

In a full analysis this step replaces the loop over SMMAT results:

```{r smmat-combine, eval=FALSE}
# Collect p_SMMAT from each ancestry into a matrix (genes × pops)
p_mat <- do.call(cbind, lapply(smmat_results, `[[`, "p_SMMAT"))
rownames(p_mat) <- smmat_results[[1]]$gene

# Apply Cauchy combination across populations for each gene
p_cauchy_genes <- apply(p_mat, 1, cauchy_combine)
results <- data.frame(
  gene     = rownames(p_mat),
  as.data.frame(p_mat),
  p_cauchy = p_cauchy_genes
)

print(results[order(results$p_cauchy), ])
```

A gene is flagged as **ancestry-specific** when `p_cauchy` is significant but
the ancestry-specific p-values differ strongly — e.g. `p_afr << p_eur` with
`p_cauchy` driven primarily by the AFR signal.

---

## 8  Cleanup

```{r cleanup}
unlink(tmp, recursive = TRUE)
```

---

## Quick reference

| Step | Function | Package |
|------|----------|---------|
| Split genotypes | `run_combined_pipeline()` | lantern |
| Write dosage GDS | `write_dosage_gds()` | lantern |
| Fit null model | `glmmkin()` | GMMAT |
| Run burden test | `SMMAT()` | GMMAT |
| Combine p-values | `cauchy_combine()` | lantern |

For the two-population unphased workflow (no phased VCF required) see
`vignette("vcf-splitting")`.
For K > 2 ancestries with phased data see `vignette("multi-ancestry-splitting")`.
