---
title: "Complete LANTERN workflow: JHS chr19 real-data example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Complete LANTERN workflow: JHS chr19 real-data example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(lantern)
library(SeqArray)
RAW      <- "../../raw"
HAS_DATA <- file.exists(file.path(RAW, "LANTERN.phased.bcf"))
HAS_GDS  <- file.exists(file.path(RAW, "chr19_AFR_phased.gds"))
HAS_RDS  <- file.exists(file.path(RAW, "chr19_combined_split.rds"))
HAS_SMMAT_CACHE    <- file.exists(file.path(RAW, "smmat_AFR_phased.rds"))
HAS_SMMAT_UNPHASED <- file.exists(file.path(RAW, "smmat_AFR_unphased.rds"))
HAS_GMMAT <- requireNamespace("GMMAT", quietly = TRUE)
```

## Overview

This vignette demonstrates the complete LANTERN pipeline on real genotype data
from the Jackson Heart Study (JHS) cohort — chromosome 19, ~3 300 samples.
The same code applies to any cohort with a phased VCF and an RFMix MSP file.

```
Phased VCF/BCF  +  RFMix MSP
          │
          ▼
  run_combined_pipeline()         ← ~3 min on chr19 / 3 313 samples
          │
   ┌──────┴──────┐
phased split  unphased split     ← per-population dosage matrices
   │
write_dosage_gds()               ← one GDS per population (~2 min each)
   │
glmmkin()  ← phenotype + kinship  ← once per analysis (~9 s)
   │
SMMAT()    ← per population, per gene
   │
cauchy_combine()                 ← single multi-ancestry p-value per gene
```

The first three steps are run locally and their outputs are loaded from
cached RDS / GDS files on subsequent runs.  GMMAT-dependent steps are shown
with expected output; they require GMMAT ≥ 1.4 (not on CRAN — install from
the GMMAT GitHub or Bioconductor).

---

## 1  Split genotypes by local ancestry

`run_combined_pipeline()` parses the BCF and the RFMix MSP file once and
returns ancestry-stratified dosage matrices for both the phased and unphased
splitting methods.  The phased split is deterministic (each haplotype's dosage
goes to the population of that haplotype); the unphased split is proportional
(mixed singletons get equal weight to each ancestry).

```{r split, eval=HAS_DATA}
# Run the combined pipeline; cache to RDS on first run
cache_rds <- file.path(RAW, "chr19_combined_split.rds")

if (file.exists(cache_rds)) {
  combined <- readRDS(cache_rds)
} else {
  combined <- run_combined_pipeline(
    vcf_path = file.path(RAW, "LANTERN.phased.bcf"),
    msp_path = file.path(RAW, "LANTERN_chr19.msp.tsv.gz"),
    chrom    = "chr19",
    verbose  = TRUE
  )
  saveRDS(combined, cache_rds)
}
```

```{r split-summary, eval=HAS_RDS}
combined  <- readRDS(file.path(RAW, "chr19_combined_split.rds"))
pop_names <- setdiff(names(combined$phased),
                     c("variant_info", "sample_ids", "overlap"))
ov        <- combined$phased$overlap

cat("Populations:            ", paste(pop_names, collapse = ", "), "\n")
cat("Common samples:         ", ov$n_common, "\n")
cat("Variants (raw):         ", ov$n_variants_kept + ov$n_monomorphic_filtered +
                                ov$n_no_tract, "\n")
cat("  No-tract dropped:     ", ov$n_no_tract, "\n")
cat("  Monomorphic filtered: ", ov$n_monomorphic_filtered, "\n")
cat("  Kept for analysis:    ", ov$n_variants_kept, "\n")
```

```{r split-dosage-summary, eval=HAS_RDS}
vi   <- combined$phased$variant_info
ids  <- combined$phased$sample_ids

cat("Chromosome range: ", vi$chrom[1], min(vi$pos), "–", max(vi$pos), "\n\n")

for (p in pop_names) {
  ds_ph <- combined$phased[[p]]
  ds_un <- combined$unphased[[p]]
  n_alt  <- sum(rowSums(ds_ph) > 0)
  n_diff <- sum(abs(ds_ph - ds_un) > 1e-9)   # phased ≠ unphased (mixed carriers)
  cat(sprintf("%-3s phased:   %5d variants with alt carriers  mean dosage = %.5f\n",
              p, n_alt, mean(ds_ph)))
  cat(sprintf("%-3s unphased: cells phased≠unphased = %d  (%.2f%% of all cells)\n\n",
              p, n_diff, 100 * n_diff / (nrow(ds_ph) * ncol(ds_ph))))
}
```

The table above shows that 39 089 of the 44 271 variants have at least one
AFR-ancestry carrier (vs. 16 236 EUR-ancestry carriers), reflecting the
predominantly African-American ancestry of the JHS cohort (median per-sample
AFR dosage fraction ≈ 87 %).

Phased and unphased dosages differ in only ~0.03 % of cells — those where a
mixed-ancestry carrier carries exactly one alt allele and phased data assigns
it to a specific haplotype, while the unphased split gives 0.5 to each
ancestry.

---

## 2  Per-sample ancestry proportion

```{r ancestry-fraction, eval=HAS_RDS}
afr_sum  <- colSums(combined$phased$AFR)
eur_sum  <- colSums(combined$phased$EUR)
frac_afr <- afr_sum / (afr_sum + eur_sum + .Machine$double.eps)

cat("Per-sample AFR dosage fraction (quantiles):\n")
print(round(quantile(frac_afr, c(0.05, 0.25, 0.50, 0.75, 0.95)), 3))
```

The quantiles confirm the JHS is an African-American cohort: even the 5th
percentile is ~50 % AFR ancestry.  LANTERN's ancestry-stratified approach
is most powerful in precisely these admixed populations where the same
genomic region carries distinct rare-variant spectra from each contributing
ancestral population.

---

## 3  Write ancestry-specific GDS files

`write_dosage_gds()` converts each ancestry's dosage matrix to a SeqArray
GDS file for downstream use by SMMAT.  On first run each GDS takes roughly
2 minutes for 44 271 variants × 3 313 samples; subsequent runs load from disk.

```{r write-gds, eval=HAS_RDS}
vi  <- combined$phased$variant_info
ids <- combined$phased$sample_ids

for (mode in c("phased", "unphased")) {
  for (pop in pop_names) {
    gp <- file.path(RAW, paste0("chr19_", pop, "_", mode, ".gds"))
    if (!file.exists(gp)) {
      cat("Writing", basename(gp), "...\n")
      write_dosage_gds(combined[[mode]][[pop]], vi, ids, gp)
    } else {
      cat("Cached:", basename(gp), "(", round(file.size(gp)/1e6, 1), "MB )\n")
    }
  }
}
```

```{r verify-gds, eval=HAS_GDS}
gds <- seqOpen(file.path(RAW, "chr19_AFR_phased.gds"))
sm  <- seqSummary(gds, check = "none", verbose = FALSE)
cat("AFR phased GDS:  variants =", sm$num.variant,
    "  samples =", sm$num.sample, "\n")
seqClose(gds)
```

---

## 4  Gene-group file

SMMAT reads variant–gene membership from a tab-delimited file with no header
and columns: **gene, chrom, pos, ref, alt, weight**.

The gene coordinates below come from Ensembl annotations for chr19 genes that
have ≥ 5 rare-variant entries in the JHS VCF (1 442 genes, 50 240 variant
entries after matching by position/allele):

```{r gene-group, eval=HAS_GDS}
gene_file <- file.path(RAW, "genes_chr19_smmat.tsv")
genes     <- read.table(gene_file, header = FALSE, sep = "\t",
               col.names = c("gene", "chrom", "pos", "ref", "alt", "weight"))

cat("Genes:    ", length(unique(genes$gene)), "\n")
cat("Variants: ", nrow(genes), "\n")
cat("Chrom:    ", unique(genes$chrom), "\n\n")
cat("First gene (A1BG):\n")
print(head(genes[genes$gene == "A1BG", ], 4))
```

> **Note**: the chromosome column must match what SeqArray stores in the GDS
> (here `"19"`, without the `"chr"` prefix, even though the source BCF uses
> `"chr19"`).

---

## 5  Null model  *(requires GMMAT)*

The GLMM null model is fit **once** and shared across all SMMAT calls.
For a real JHS analysis you would use the actual kinship / GRM matrix and
observed phenotype; here we illustrate with the supplied relatedness matrix
and a simulated binary phenotype.

```{r null-model, eval=HAS_GMMAT && HAS_RDS}
library(GMMAT)

combined <- readRDS(file.path(RAW, "chr19_combined_split.rds"))
ids <- combined$phased$sample_ids
K   <- readRDS(file.path(RAW, "LANTERN_kinship_rds.rds"))
K   <- K[ids, ids]   # subset to common samples in pipeline order

set.seed(123)
df_pheno <- data.frame(id = ids,
                       y  = rbinom(length(ids), 1L, 0.3))

cat("Fitting null GLMM (N =", length(ids), ")...\n")
t0     <- proc.time()
model0 <- glmmkin(
  y ~ 1,
  data   = df_pheno,
  kins   = K,
  id     = "id",
  family = binomial(link = "logit")
)
cat("Done in", round((proc.time() - t0)["elapsed"], 1), "s\n")
```

> For a **quantitative** trait use `family = gaussian(link = "identity")`.
> For **case-control** data with many cases use `method.optim = "Brent"` in
> `glmmkin()` to speed up variance-component estimation.

---

## 6  SMMAT gene-level tests  *(requires GMMAT)*

One SMMAT call per population × split mode.  Each call processes all 1 442
genes in one pass and takes roughly 5–8 minutes on 1 core.

```{r smmat-run, eval=HAS_GMMAT && HAS_RDS}
gene_file <- file.path(RAW, "genes_chr19_smmat.tsv")

run_smmat <- function(gds_file) {
  SMMAT(
    null.obj   = model0,
    geno.file  = gds_file,
    group.file = gene_file,
    MAF.range  = c(0, 0.05),    # rare variants only (MAF < 5 %)
    miss.cutoff = 1,
    method     = "davies",
    is.dosage  = TRUE,          # required for ancestry-specific dosage GDS
    ncores     = 1L,
    verbose    = FALSE
  )
}

smmat_results <- list()
for (mode in c("phased", "unphased")) {
  for (pop in pop_names) {
    key   <- paste0(pop, "_", mode)
    cache <- file.path(RAW, paste0("smmat_", pop, "_", mode, ".rds"))
    if (file.exists(cache)) {
      smmat_results[[key]] <- readRDS(cache)
      cat("Loaded:", cache, "\n")
    } else {
      cat("Running SMMAT", key, "...\n")
      t0 <- proc.time()
      smmat_results[[key]] <- run_smmat(
        file.path(RAW, paste0("chr19_", pop, "_", mode, ".gds")))
      saveRDS(smmat_results[[key]], cache)
      cat("  done in", round((proc.time() - t0)["elapsed"] / 60, 1), "min\n")
    }
  }
}
```

Each result is a data frame with one row per gene.  Key columns:

| Column | Meaning |
|--------|---------|
| `group` | gene name |
| `n.variants` | rare variants included |
| `freq.mean` | mean alt-allele frequency across included variants |
| `B.score` / `B.pval` | burden (score) test statistic / p-value |
| `E.pval` | efficient omnibus p-value (primary SMMAT result) |

---

## 7  Cauchy combination across ancestries

`cauchy_combine()` pools the per-ancestry SMMAT p-values into a single
multi-ancestry gene p-value.  It is valid under the arbitrary correlation
that exists between AFR and EUR dosage matrices within an admixed cohort.

```{r cauchy-combine, eval=HAS_SMMAT_CACHE}
# Load phased SMMAT results
res_afr <- readRDS(file.path(RAW, "smmat_AFR_phased.rds"))
res_eur <- readRDS(file.path(RAW, "smmat_EUR_phased.rds"))

# Align by gene name
common_genes <- intersect(res_afr$group, res_eur$group)
afr <- res_afr[match(common_genes, res_afr$group), ]
eur <- res_eur[match(common_genes, res_eur$group), ]

results <- data.frame(
  gene      = afr$group,
  n_var_afr = afr$n.variants,
  n_var_eur = eur$n.variants,
  p_AFR     = afr$E.pval,
  p_EUR     = eur$E.pval,
  stringsAsFactors = FALSE
)

# Apply Cauchy combination where both ancestries have results
both <- !is.na(results$p_AFR) & !is.na(results$p_EUR) &
         results$p_AFR > 0     & results$p_EUR > 0
results$p_cauchy <- NA_real_
results$p_cauchy[both] <- mapply(
  function(a, e) cauchy_combine(c(a, e)),
  results$p_AFR[both], results$p_EUR[both]
)

cat("Genes analysed:", nrow(results), "\n")
cat("Genes with both AFR and EUR p-values:", sum(both), "\n\n")

cat("Top 15 genes by Cauchy p-value (simulated phenotype):\n")
print(head(results[order(results$p_cauchy), ], 15))
```

```{r cauchy-plot, eval=HAS_SMMAT_CACHE, fig.width=6, fig.height=5}
# Compare ancestry-specific vs. combined p-values
log_afr    <- -log10(results$p_AFR[both])
log_eur    <- -log10(results$p_EUR[both])
log_cauchy <- -log10(results$p_cauchy[both])

plot(log_afr, log_cauchy,
     xlab = expression(-log[10](p[AFR])),
     ylab = expression(-log[10](p[Cauchy])),
     pch  = 16, cex = 0.5, col = adjustcolor("steelblue", 0.5),
     main = "Cauchy vs. AFR-only gene p-value (phased split, chr19)")
abline(0, 1, lty = 2, col = "grey50")

# Highlight genes where EUR is stronger (EUR > 2× AFR on log scale)
eu_driven <- log_eur > log_afr + 2 & log_eur > 2
points(log_afr[eu_driven], log_cauchy[eu_driven],
       pch = 16, cex = 0.7, col = "tomato")
legend("topleft",
       c("EUR-driven signal", "reference (y = x)"),
       pch = c(16, NA), lty = c(NA, 2),
       col = c("tomato", "grey50"), bty = "n")
```

```{r phased-vs-unphased, eval=HAS_SMMAT_CACHE && HAS_SMMAT_UNPHASED, fig.width=6, fig.height=5}
# Compare phased vs. unphased Cauchy p-values
res_afr_un <- readRDS(file.path(RAW, "smmat_AFR_unphased.rds"))
res_eur_un <- readRDS(file.path(RAW, "smmat_EUR_unphased.rds"))

afr_un <- res_afr_un[match(common_genes, res_afr_un$group), ]
eur_un <- res_eur_un[match(common_genes, res_eur_un$group), ]

both_un <- !is.na(afr_un$E.pval) & !is.na(eur_un$E.pval) &
            afr_un$E.pval > 0    & eur_un$E.pval > 0

p_cauchy_un <- rep(NA_real_, nrow(results))
p_cauchy_un[both_un] <- mapply(
  function(a, e) cauchy_combine(c(a, e)),
  afr_un$E.pval[both_un], eur_un$E.pval[both_un]
)

# Scatter: phased Cauchy vs. unphased Cauchy
keep <- both & both_un
plot(-log10(results$p_cauchy[keep]), -log10(p_cauchy_un[keep]),
     xlab = expression(-log[10](p[Cauchy~phased])),
     ylab = expression(-log[10](p[Cauchy~unphased])),
     pch  = 16, cex = 0.5, col = adjustcolor("darkgreen", 0.4),
     main = "Phased vs. unphased Cauchy p-value (chr19)")
abline(0, 1, lty = 2, col = "grey50")
```

The phased and unphased Cauchy p-values are nearly identical for this cohort
(points cluster along the diagonal).  The two methods differ only for variants
where a mixed-ancestry carrier holds the alt allele on a single haplotype —
fewer than 0.03 % of cells in the JHS chr19 data.  In practice, use the
phased split whenever high-quality phased data are available.

---

## 8  Quick-reference: key numbers from this run

```{r summary-table, eval=HAS_SMMAT_CACHE && HAS_RDS}
cat(sprintf("%-35s %s\n", "Metric", "Value"))
cat(strrep("-", 50), "\n")
ov <- combined$phased$overlap
cat(sprintf("%-35s %d\n",  "Samples analysed",                ov$n_common))
cat(sprintf("%-35s %d\n",  "Variants analysed",               ov$n_variants_kept))
cat(sprintf("%-35s %d\n",  "Variants with AFR carrier",
            sum(rowSums(combined$phased$AFR) > 0)))
cat(sprintf("%-35s %d\n",  "Variants with EUR carrier",
            sum(rowSums(combined$phased$EUR) > 0)))
cat(sprintf("%-35s %d\n",  "Genes tested",                    nrow(results)))
cat(sprintf("%-35s %d\n",  "Genes with both AFR+EUR p-values", sum(both)))
cat(sprintf("%-35s %.3g\n","Min Cauchy p-value",              min(results$p_cauchy, na.rm=TRUE)))
```

---

## 9  Function reference

| Step | Function | Package |
|------|----------|---------|
| Split genotypes | `run_combined_pipeline()` | lantern |
| Write dosage GDS | `write_dosage_gds()` | lantern |
| Fit null GLMM | `glmmkin()` | GMMAT |
| Gene burden test | `SMMAT()` | GMMAT |
| Combine p-values | `cauchy_combine()` | lantern |

For the two-population unphased workflow see `vignette("vcf-splitting")`.
For K > 2 ancestries see `vignette("multi-ancestry-splitting")`.
For the unphased-only workflow starting from dosage matrices see
`vignette("gene-level-test")`.
