---
title: "Ancestry-stratified gene-level burden tests"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ancestry-stratified gene-level burden tests}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(lantern)
```

## Overview

After splitting genotype dosages by local ancestry (see
`vignette("vcf-splitting")`), LANTERN feeds the resulting ancestry-specific
dosage matrices into **SMMAT** (Sequence Kernel Association Mixed Model
Association Test) via the **GMMAT** package.  Three parallel tests are run
for each gene:

| Label | Dosage input | Captures |
|-------|--------------|---------|
| **AA** | African-ancestry dosage | effects acting through AFR haplotypes |
| **EE** | European-ancestry dosage | effects acting through EUR haplotypes |
| **observed** | Raw unphased dosage | aggregate signal (serves as baseline) |

The AA and EE p-values can then be combined with a Cauchy or Fisher
meta-analysis step to produce a single multi-ancestry gene p-value.

---

## 1  Toy data setup

We simulate a small dataset that mirrors the real-data workflow: 60
independent samples, 1 gene with 18 rare variants.

```{r toy-setup}
set.seed(2025)
N      <- 60     # samples
N_VAR  <- 18     # variants in the toy gene
GENE   <- "TOYGENE1"
CHROM  <- "1"

# Simulate independent samples (identity kinship) —
# in real data replace with the LANTERN_kinship_rds.rds matrix
Kmat    <- diag(N)
samp_id <- paste0("S", sprintf("%03d", seq_len(N)))
rownames(Kmat) <- colnames(Kmat) <- samp_id

# Binary phenotype (case/control) with case fraction ~30%
pheno <- rbinom(N, 1, 0.3)
df_pheno <- data.frame(id = samp_id, y = pheno)
```

### 1.1  Simulate ancestry-split dosage matrices

We start from a raw unphased GT matrix and a per-variant ancestry code
matrix, then call the LANTERN C backend.

```{r simulate-dosages}
# Rare variant dosages: mostly 0, occasionally 1, rarely 2
gt_raw <- matrix(
  sample(0:2, N_VAR * N, replace = TRUE, prob = c(0.90, 0.09, 0.01)),
  nrow = N_VAR, ncol = N
)
rownames(gt_raw) <- paste0(CHROM, ":", seq(1e6, by = 500, length.out = N_VAR))
colnames(gt_raw) <- samp_id

# Ancestry codes: 1=EUR, 2=MIX, 3=AFR — admixed cohort with ~40% AFR
anc_raw <- matrix(
  sample(1:3, N_VAR * N, replace = TRUE, prob = c(0.35, 0.25, 0.40)),
  nrow = N_VAR, ncol = N,
  dimnames = dimnames(gt_raw)
)

split_out <- split_by_ancestry(gt_raw, anc_raw)
ds_afr <- split_out$african    # variants × samples, African ancestry dosage
ds_eur <- split_out$european   # variants × samples, European ancestry dosage

cat("AFR dosage: ", nrow(ds_afr), "variants ×", ncol(ds_afr), "samples\n")
cat("EUR dosage: ", nrow(ds_eur), "variants ×", ncol(ds_eur), "samples\n")
cat("AFR mean per-variant dosage:", round(mean(ds_afr), 4), "\n")
cat("EUR mean per-variant dosage:", round(mean(ds_eur), 4), "\n")
```

---

## 2  Packaging dosage matrices into GDS files

SMMAT reads variants from **SeqArray GDS** files with a `DS` (dosage) FORMAT
field.  We create them from temporary VCFs.

```{r create-gds, message=FALSE}
library(SeqArray)

# Variant metadata matching our simulated positions
positions  <- seq(1000000L, by = 500L, length.out = N_VAR)
ref_allele <- rep("A", N_VAR)
alt_allele <- rep("T", N_VAR)

vi <- data.frame(
  chrom = CHROM, pos = positions,
  ref   = ref_allele, alt = alt_allele,
  stringsAsFactors = FALSE
)

gds_afr <- tempfile(fileext = ".gds")
gds_eur <- tempfile(fileext = ".gds")

# write_dosage_gds() handles the VCF → GDS conversion
write_dosage_gds(ds_afr, vi, samp_id, gds_afr)
write_dosage_gds(ds_eur, vi, samp_id, gds_eur)

cat("GDS files created:\n", gds_afr, "\n", gds_eur, "\n")
```

---

## 3  Gene-group file

SMMAT looks up variants per gene from a tab-delimited file with no header:

```
gene_name   chrom   pos   ref   alt   weight
```

The `weight` column is typically 1 (unweighted burden) or a function of MAF
(e.g. Beta(1,25) weights).

```{r gene-group}
gene_group_file <- tempfile(fileext = ".tsv")
gene_df <- data.frame(
  gene  = GENE,
  chr   = CHROM,
  pos   = positions,
  ref   = ref_allele,
  alt   = alt_allele,
  weight = 1
)
write.table(gene_df, gene_group_file,
            sep = "\t", row.names = FALSE, col.names = FALSE, quote = FALSE)

cat("Gene-group file (first 3 rows):\n")
head(read.table(gene_group_file, sep = "\t"), 3)
```

---

## 4  Fitting the null model

The null mixed model is fit **once** and reused for all three SMMAT calls:

```{r null-model, eval=FALSE}
library(GMMAT)

# glmmkin: logistic GLMM for case/control phenotype
# For a continuous trait use family = gaussian(link = "identity")
model0 <- glmmkin(
  y ~ 1,
  data   = df_pheno,
  kins   = Kmat,
  id     = "id",
  family = binomial(link = "logit")
)
```

> **Note**: GMMAT >= 1.4 is required.  In real analyses the kinship matrix
> (`kins`) is a sparse relatedness matrix estimated from genotype data; here
> we pass the identity matrix to assume independence.

---

## 5  Running SMMAT gene-level tests

```{r smmat-run, eval=FALSE}
run_smmat <- function(gds_file, is_dosage) {
  SMMAT(
    model0,
    gds.fn      = gds_file,
    group.file  = gene_group_file,
    MAF.range   = c(0, 0.05),  # rare variants only (MAF < 5%)
    miss.cutoff = 1,            # keep all variants
    method      = "davies",     # numerical integration for p-values
    is.dosage   = is_dosage,    # TRUE for ancestry-specific dosage
    ncores      = 1L
  )
}

res_aa  <- run_smmat(gds_afr, is_dosage = TRUE)
res_ee  <- run_smmat(gds_eur, is_dosage = TRUE)
res_obs <- run_smmat(gds_obs, is_dosage = FALSE)  # raw unphased GT
```

Each result is a data.frame with one row per gene and columns:
`gene`, `N.variant`, `N.sample`, `B`, `B.var`, `Q_B`, `p_B`, `Q_S`, `p_S`,
`p_SMMAT` (the primary omnibus p-value).

---

## 6  Interpreting and combining results

```{r smmat-combine, eval=FALSE}
# Bind AA / EE / observed into a single data.frame for the gene
combined <- data.frame(
  gene  = res_aa$gene,
  p_aa  = res_aa$p_SMMAT,
  p_ee  = res_ee$p_SMMAT,
  p_obs = res_obs$p_SMMAT
)

# Cauchy combination of AA and EE p-values (see ?cauchy_combine)
combined$p_cauchy <- mapply(function(a, e) cauchy_combine(c(a, e)),
                             combined$p_aa, combined$p_ee)

print(combined)
```

A significant `p_cauchy` with `p_aa` and/or `p_ee` more extreme than
`p_obs` suggests the signal is ancestry-specific — the key LANTERN finding.

---

## 7  Cleanup

```{r cleanup}
unlink(c(gds_afr, gds_eur, gene_group_file))
```

---

## 8  End-to-end workflow diagram

```
Phased VCF + RFMix MSP
         │
         ▼
  run_phased_pipeline()
         │
    ┌────┴────┐
    │         │
african_ds.gds  european_ds.gds
    │         │
    └────┬────┘
         │
    glmmkin()  ← phenotype + kinship
         │
    ┌────┼────┐
    │    │    │
 SMMAT SMMAT SMMAT
  AA    EE   obs
    │    │
  Cauchy combination
         │
  gene-level p-value
```

See `vignette("vcf-splitting")` for the first step of this workflow.
