---
title: "Getting started with lantern"
author:
  - name: Yu Wang
  - name: Bjoernar Tuftin
  - name: Paul Auer
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Getting started with lantern}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Introduction

In admixed individuals, each stretch of the genome is inherited from one of
several ancestral populations. A rare-variant burden test that pools all
carriers can miss an effect that exists in only one ancestral background,
or wrongly attribute one. **lantern** uses local ancestry to split each
sample's alternate-allele dosage into per-ancestry dosages (e.g. African
and European). It then runs gene-level `SMMAT` tests
([GMMAT](https://cran.r-project.org/package=GMMAT)) separately for each
ancestry and combines the per-ancestry p-values with the Cauchy combination
test.

The workflow has three steps:

1. **Split** a phased VCF by local ancestry from an RFMix `.msp.tsv` file
   with `ancestry_split()`.
2. **Write** each ancestry's dosages to a `r Biocpkg("SeqArray")` GDS file
   with `write_ancestry_gds()`.
3. **Test** each gene in every ancestry and Cauchy-combine the results with
   `ancestry_smmat()`.

# Installation

```{r install, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("lantern")
```

```{r load}
library(lantern)
```

lantern reads VCFs with `bcftools` when it is on `PATH`, which is fastest
and also reads BCF. Otherwise it falls back to `r Biocpkg("SeqArray")`, so
no external tools are needed.

# Example data

The package ships a small simulated dataset: 60 admixed samples, 200 rare
variants on chromosome 19, two ancestral populations (AFR, EUR) and 8 genes.
The phenotype carries an effect that is driven only by African-ancestry
alleles in `GENE2`.

```{r files}
ext <- system.file("extdata", package = "lantern")
vcf <- file.path(ext, "toy.vcf.gz")     # phased genotypes
msp <- file.path(ext, "toy.msp.tsv")    # RFMix local ancestry
substr(readLines(msp, n = 3), 1, 70)
```

# Step 1: split dosages by local ancestry

`mode = "haplotype"` assigns each phased allele to the ancestry of its own
haplotype, so it is exact. `mode = "dosage"` ignores phase: it splits
ambiguous mixed-ancestry heterozygotes using population proportions
estimated from the unambiguous carriers, shrunk toward each chromosome
arm's global ancestry. See `vignette("split-intuition")` for the details.

```{r split}
split <- ancestry_split(vcf, msp, mode = "dosage", chrom = "chr19",
                        verbose = FALSE)
names(split)
dim(split$AFR)
split$AFR[1:4, 1:6]
split$EUR[1:4, 1:6]
```

The result also records which variants and samples were kept:

```{r kept}
head(split$variant_info)
split$overlap[c("n_common", "n_variants_kept", "n_monomorphic_filtered")]
```

# Step 2: write per-ancestry GDS files

```{r gds}
out_dir <- tempfile("lantern-")
gds <- write_ancestry_gds(split, out_dir, verbose = FALSE)
gds
```

# Step 3: ancestry-stratified gene tests

`ancestry_smmat()` fits the null model once with `GMMAT::glmmkin()`, then
runs `GMMAT::SMMAT()` on each ancestry's GDS file. Gene-level weights for
the Cauchy combination are the median per-variant carrier counts in each
ancestry. The samples here are unrelated, so the kinship matrix is the
identity.

```{r smmat, eval = requireNamespace("GMMAT", quietly = TRUE) && requireNamespace("SeqVarTools", quietly = TRUE)}
pheno <- read.delim(file.path(ext, "toy_pheno.tsv"))
kinship <- diag(nrow(pheno))
dimnames(kinship) <- list(pheno$id, pheno$id)

res <- ancestry_smmat(gds, pheno, y ~ age + sex, kinship,
                      gene_group_file = file.path(ext, "toy_genes.tsv"),
                      ancestry_counts = split$ancestry_counts,
                      variant_info    = split$variant_info,
                      verbose = FALSE)
res$results
```

`GENE2` is strongly associated in the AFR test but not in the EUR test,
which matches the simulated African-specific effect. A test that pools both
ancestries would dilute that signal.

```{r cleanup, include = FALSE}
unlink(out_dir, recursive = TRUE)
```

# Further reading

- `vignette("split-intuition")`: what the split computes, on small
  matrices, for two or more populations.
- `vignette("toy-vcf-msp-example")`: the VCF and MSP input formats, with
  four populations.
- `vignette("real-data-chr19")`: a real-cohort analysis with binary,
  continuous and count phenotypes.

# Session information

```{r session}
sessionInfo()
```
