---
title: "Toy example: splitting a VCF + MSP file (two and more populations)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Toy example: splitting a VCF + MSP file (two and more populations)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(lantern)
```

## Overview

`ancestry_split()` (Step 1) is LANTERN's entry point once you have a
phased VCF and an RFMix MSP file: it parses both, intersects samples, maps
variants to ancestry tracts, and runs either split via a single `mode`
argument (`"haplotype"` or `"dosage"`) — see `vignette("split-intuition")`
for what each mode computes on the underlying matrices.

Population count and names come directly from the MSP header, so the same
call works unchanged for K = 2 or K > 2. Part A builds a two-population
toy VCF + MSP; Part B does the same for four populations.

---

# Part A: Two populations

```{r toy-files}
tmp_dir  <- tempfile()
dir.create(tmp_dir)
vcf_path <- file.path(tmp_dir, "phased.vcf")
msp_path <- file.path(tmp_dir, "ancestry.msp.tsv")

# 3 variants × 3 samples, phased GTs
writeLines(c(
  "##fileformat=VCFv4.1",
  '##FORMAT=<ID=GT,Number=1,Type=String,Description="Genotype">',
  paste(c("#CHROM","POS","ID","REF","ALT","QUAL","FILTER","INFO",
          "FORMAT","S1","S2","S3"), collapse = "\t"),
  paste(c("19","1100",".","A","G","100","PASS",".","GT","0|1","1|0","0|0"), collapse = "\t"),
  paste(c("19","1200",".","C","T","100","PASS",".","GT","1|1","0|1","1|0"), collapse = "\t"),
  paste(c("19","2100",".","G","A","100","PASS",".","GT","0|0","1|0","0|1"), collapse = "\t")
), vcf_path)

# RFMix MSP: 2 ancestry tracts, pop codes AFR=0 EUR=1
#   tract 1 [1000,2000): S1=AFR|EUR  S2=EUR|AFR  S3=AFR|EUR
#   tract 2 [2000,3000): S1=EUR|AFR  S2=AFR|EUR  S3=EUR|AFR
writeLines(c(
  paste(c("#Subpopulation order/codes: AFR=0", "EUR=1"), collapse = "\t"),
  paste(c("#chm","spos","epos","sgpos","egpos","n snps",
          "S1.0","S1.1","S2.0","S2.1","S3.0","S3.1"), collapse = "\t"),
  paste(c("19","1000","2000","0.0","0.1","2","0","1","1","0","0","1"), collapse = "\t"),
  paste(c("19","2000","3000","0.1","0.2","1","1","0","0","1","1","0"), collapse = "\t")
), msp_path)
```

```{r toy-split}
phased   <- ancestry_split(vcf_path, msp_path, mode = "haplotype", verbose = FALSE)
unphased <- ancestry_split(vcf_path, msp_path, mode = "dosage",    verbose = FALSE)

pop_names <- setdiff(names(phased),
                     c("variant_info", "sample_ids", "mode",
                       "ancestry_counts", "tract_info", "overlap"))
cat("Populations:", paste(pop_names, collapse = ", "), "\n\n")

cat("=== phased (mode = \"haplotype\") ===\n")
for (p in pop_names) { cat(p, "dosage:\n"); print(phased[[p]]) }

cat("\n=== unphased (mode = \"dosage\") ===\n")
for (p in pop_names) { cat(p, "dosage:\n"); print(unphased[[p]]) }

cat("\nVariants kept:", nrow(phased$variant_info), "\n")
cat("Samples:      ", length(phased$sample_ids), "\n")

# Sanity check: phased AFR + EUR reconstructs the raw diploid genotype exactly
vcf_lines <- readLines(vcf_path)[-(1:3)]
raw <- do.call(rbind, lapply(vcf_lines, function(l) {
  gts <- strsplit(l, "\t")[[1]][10:12]
  vapply(gts, function(g) sum(as.integer(strsplit(g, "|", fixed = TRUE)[[1]])), integer(1))
}))
max_err <- max(abs((phased$AFR + phased$EUR) - raw))
cat("Max phased reconstruction error (must be 0):", max_err, "\n")
```

Each call returns a named list: one dosage matrix per population (named
from the MSP header) plus `variant_info`, `sample_ids`, `mode`,
`ancestry_counts`, and `overlap`. Need both splits on a large cohort
routinely? Cache each result (e.g. to RDS) rather than re-parsing the
VCF/MSP on every run — `ancestry_split()` always re-parses from scratch.

The backward-compatible `ancestry_split_dosage()` (matrix or VCF/MSP
input) and `ancestry_split_phased()` (VCF/MSP input, optionally also
writing GDS) wrap the same logic for existing code — see
`?ancestry_split_dosage` and `?ancestry_split_phased`.

---

# Part B: More than two populations (K = 4)

Nothing about the call changes — same function, same `mode` argument,
same return shape, just more populations declared in the MSP header.

```{r k4-setup}
tmp_dir4 <- tempfile()
dir.create(tmp_dir4)
vcf4_path <- file.path(tmp_dir4, "phased4.vcf")
msp4_path <- file.path(tmp_dir4, "ancestry4.msp.tsv")

set.seed(99)
n_samp_k4 <- 24
samp_k4   <- paste0("S", seq_len(n_samp_k4))

# 10 variants, phased GT, mostly rare
gt_lines <- vapply(seq_len(10), function(i) {
  gts <- sample(c("0|0","0|1","1|0","1|1"), n_samp_k4, replace = TRUE,
                prob = c(0.85, 0.06, 0.06, 0.03))
  paste(c("19", 1000L + i * 100L, ".", "A", "T", "100", "PASS", ".", "GT", gts),
        collapse = "\t")
}, character(1))
writeLines(c(
  "##fileformat=VCFv4.1",
  '##FORMAT=<ID=GT,Number=1,Type=String,Description="Genotype">',
  paste(c("#CHROM","POS","ID","REF","ALT","QUAL","FILTER","INFO","FORMAT", samp_k4),
        collapse = "\t"),
  gt_lines
), vcf4_path)

# Single tract spanning all variants; each sample gets a random pair of
# the four ancestries per haplotype
pop_codes_k4 <- c(AFR = 0L, EUR = 1L, NAT = 2L, EAS = 3L)
hap_cols <- unlist(lapply(samp_k4, function(s) paste0(s, c(".0", ".1"))))
hap_anc  <- sample(pop_codes_k4, length(hap_cols), replace = TRUE)
writeLines(c(
  paste(c("#Subpopulation order/codes: AFR=0", "EUR=1", "NAT=2", "EAS=3"),
        collapse = "\t"),
  paste(c("#chm","spos","epos","sgpos","egpos","n snps", hap_cols), collapse = "\t"),
  paste(c("19", "1000", "3000", "0.0", "0.5", "10", hap_anc), collapse = "\t")
), msp4_path)
```

```{r k4-split}
phased4   <- ancestry_split(vcf4_path, msp4_path, mode = "haplotype",
                             chrom = "19", verbose = FALSE)
unphased4 <- ancestry_split(vcf4_path, msp4_path, mode = "dosage",
                             chrom = "19", verbose = FALSE)

pop4 <- setdiff(names(phased4),
                c("variant_info", "sample_ids", "mode",
                  "ancestry_counts", "tract_info", "overlap"))
cat("Populations recovered from the MSP header:", paste(pop4, collapse = ", "), "\n")
cat("Variants kept:", nrow(phased4$variant_info), "\n")
cat("ancestry_counts columns:", paste(colnames(phased4$ancestry_counts), collapse = ", "), "\n\n")

cat("=== phased AFR (first 3 variants) ===\n"); print(phased4$AFR[1:3, ])
cat("=== unphased AFR (first 3 variants) ===\n"); print(round(unphased4$AFR[1:3, ], 2))
```

Next: `vignette("real-data-chr19")` takes a split result like this one
through `write_ancestry_gds()` (Step 2) and `ancestry_smmat()` (Step 3) on
real genotype data.
