---
title: "Splitting genotypes by local ancestry"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Splitting genotypes by local ancestry}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(lantern)
```

## Overview

LANTERN splits variant dosages by local ancestry so that downstream
association tests can be run separately on the African (AA) and European
(EE) haplotype backgrounds of admixed individuals.

Two input modes are supported:

| Mode | Input | Key function |
|------|-------|-------------|
| **Unphased** | Genotype matrix + per-variant ancestry code | `split_diploid()` / `ancestry_split_dosage()` |
| **Phased** | Per-haplotype GT + per-haplotype ancestry | `split_haplotype()` / `ancestry_split_phased()` |

---

## 1  Unphased splitting

### 1.1  Anatomy of the input matrices

`split_diploid()` takes two matrices of identical dimensions:

* **`gt_genotype`** — integer matrix, rows = variants, cols = samples.
  Values: 0 (hom-ref), 1 (het), 2 (hom-alt).
* **`ancestry`** — integer matrix, same shape.
  Ancestry codes: **1** = EUR/EUR, **2** = AFR/EUR (mixed het), **3** = AFR/AFR.

The ancestry code summarises *both* haplotypes at the locus (e.g. from an
RFMix MSP file collapsed into a per-variant diploid code).

### 1.2  Minimal working example

```{r unphased-basic}
set.seed(42)
n_var  <- 10   # variants
n_samp <- 6    # samples

# Simulate integer genotypes (0/1/2)
gt <- matrix(
  sample(0:2, n_var * n_samp, replace = TRUE, prob = c(0.6, 0.3, 0.1)),
  nrow = n_var, ncol = n_samp
)
rownames(gt) <- paste0("var", seq_len(n_var))
colnames(gt) <- paste0("S", seq_len(n_samp))

# Simulate ancestry codes (1=EUR, 2=MIX, 3=AFR)
anc <- matrix(
  sample(1:3, n_var * n_samp, replace = TRUE, prob = c(0.4, 0.2, 0.4)),
  nrow = n_var, ncol = n_samp
)

result <- split_diploid(gt, anc)
```

```{r unphased-inspect}
cat("African dosage (first 4 variants × 6 samples):\n")
round(result$african[1:4, ], 2)

cat("\nEuropean dosage (first 4 variants × 6 samples):\n")
round(result$european[1:4, ], 2)
```

### 1.3  How the split works

For each variant the function computes ancestry-specific proportions
**p1** (African) and **p2** (European) from the diploid counts across all
samples:

```
p1 = (2·N_AA_hom + N_AA_het + N_MIX_hom) / denominator
p2 = (N_MIX_hom + 2·N_EUR_hom + N_EUR_het) / denominator
```

where `denominator = total_alt_alleles − N_MIX_het` (mixed-het samples
are assigned stochastically via p1/p2 rather than excluded from the
denominator).

**Singleton edge-case**: when *all* alt alleles are carried by mixed-het
individuals, p1 = p2 = 0.5 to avoid dividing by zero.

```{r unphased-edge}
# A variant where only mixed-het (code 2) samples carry the alt allele
gt_edge  <- matrix(c(1L, 0L, 1L), nrow = 1)   # three samples: het, ref, het
anc_edge <- matrix(c(2L, 1L, 2L), nrow = 1)   # all alt carriers are MIX

r <- split_diploid(gt_edge, anc_edge)
cat("Singleton MIX case — p1 = p2 = 0.5\n")
cat("African: ", r$african, "\nEuropean:", r$european, "\n")
```

### 1.4  High-level pipeline (`ancestry_split_dosage`)

`ancestry_split_dosage()` wraps the splitting in a full pipeline that
handles sample/variant overlap, monomorphic filtering, and optional VCF
output.  Its inputs are:

* **`gt_matrix`**: variant × sample genotype matrix (rownames = variant IDs,
  colnames = sample IDs).
* **`pt_matrix`**: sample × region ancestry matrix (rownames = sample IDs,
  colnames match variant IDs or genomic region strings).

```{r pipeline}
set.seed(7)
n_v <- 15; n_s <- 8

gt_full <- matrix(
  sample(0:2, n_v * n_s, replace = TRUE, prob = c(0.6, 0.3, 0.1)),
  nrow = n_v, ncol = n_s,
  dimnames = list(
    paste0("chr1:", 1000 * seq_len(n_v)),   # rownames: variant positions
    paste0("samp_", seq_len(n_s))            # colnames: sample IDs
  )
)

# PT matrix: rows = samples, cols = same variant IDs
pt_full <- matrix(
  sample(1:3, n_s * n_v, replace = TRUE, prob = c(0.35, 0.3, 0.35)),
  nrow = n_s, ncol = n_v,
  dimnames = list(
    paste0("samp_", seq_len(n_s)),
    paste0("chr1:", 1000 * seq_len(n_v))
  )
)

out <- ancestry_split_dosage(gt_full, pt_full, verbose = FALSE)

cat("African matrix: ", nrow(out$african), "variants ×",
    ncol(out$african), "samples\n")
cat("European matrix:", nrow(out$european), "variants ×",
    ncol(out$european), "samples\n")
cat("Variants kept:", out$overlap$n_variants_kept, "of",
    out$overlap$n_variants_total, "\n")
```

---

## 2  Phased splitting

When RFMix (or equivalent) provides **per-haplotype** local ancestry
calls, LANTERN can deterministically assign each haplotype's allele to the
correct ancestry background — no probabilistic splitting needed.

### 2.1  Core function: `split_haplotype()`

Inputs are four matrices of the same shape (variants × samples):

| Argument | Description |
|----------|-------------|
| `gt_hap0` | integer 0/1 allele for haplotype 0 |
| `gt_hap1` | integer 0/1 allele for haplotype 1 |
| `anc_hap0` | ancestry code for haplotype 0 |
| `anc_hap1` | ancestry code for haplotype 1 |
| `pop_codes` | named vector mapping `c(AFR=0L, EUR=1L)` (RFMix default) |

```{r phased-basic}
# 3 variants × 4 samples
gt0 <- matrix(c(1L, 0L, 1L,   # variant 1: hap0 alleles for 3 samples
                0L, 1L, 0L,   # variant 2
                1L, 1L, 0L),  # variant 3
              nrow = 3, ncol = 3)

gt1 <- matrix(c(0L, 1L, 0L,
                1L, 0L, 1L,
                0L, 1L, 1L),
              nrow = 3, ncol = 3)

# Haplotype ancestry: 0 = AFR, 1 = EUR  (RFMix convention)
a0 <- matrix(c(0L, 0L, 1L,
               1L, 0L, 0L,
               0L, 1L, 0L),
             nrow = 3, ncol = 3)

a1 <- matrix(c(1L, 1L, 0L,
               0L, 1L, 1L,
               1L, 0L, 1L),
             nrow = 3, ncol = 3)

phased_result <- split_haplotype(gt0, gt1, a0, a1)

cat("African dosage (phased):\n"); print(phased_result$african)
cat("European dosage (phased):\n"); print(phased_result$european)
```

The rule is simple: for each cell, hap0 contributes its allele to the AFR
count if `anc_hap0 == 0`, to EUR if `anc_hap0 == 1`, and nothing
otherwise.  Same for hap1.

### 2.2  Full pipeline (`ancestry_split_phased`)

`ancestry_split_phased()` orchestrates the complete workflow from raw files:

1. Parse the RFMix MSP file (ancestry tracts).
2. Query sample IDs from the VCF via `bcftools`.
3. Intersect samples between VCF and MSP.
4. Parse phased GT strings from VCF output.
5. Map each variant to its ancestry tract with `findInterval()`.
6. Broadcast tract-level ancestry to variant level.
7. Call `split_haplotype()` via the C backend.
8. Filter monomorphic variants.
9. Optionally write ancestry-specific VCFs with a DS (dosage) field
   and convert them to GDS format.

```{r phased-pipeline-setup}
# Build minimal synthetic VCF and MSP in a temp directory
tmp_dir  <- tempfile()
dir.create(tmp_dir)
vcf_path <- file.path(tmp_dir, "phased.vcf")
msp_path <- file.path(tmp_dir, "ancestry.msp.tsv")
out_path <- file.path(tmp_dir, "out")

# 3 variants × 3 samples, phased GTs
writeLines(c(
  "##fileformat=VCFv4.1",
  '##FORMAT=<ID=GT,Number=1,Type=String,Description="Genotype">',
  paste(c("#CHROM","POS","ID","REF","ALT","QUAL","FILTER","INFO",
          "FORMAT","S1","S2","S3"), collapse = "\t"),
  paste(c("19","1100",".","A","G","100","PASS",".","GT","0|1","1|0","0|0"), collapse = "\t"),
  paste(c("19","1200",".","C","T","100","PASS",".","GT","1|1","0|1","1|0"), collapse = "\t"),
  paste(c("19","2100",".","G","A","100","PASS",".","GT","0|0","1|0","0|1"), collapse = "\t")
), vcf_path)

# RFMix MSP: 2 ancestry tracts, pop codes AFR=0 EUR=1
#   tract 1 [1000,2000): S1=AFR|EUR  S2=EUR|AFR  S3=AFR|EUR
#   tract 2 [2000,3000): S1=EUR|AFR  S2=AFR|EUR  S3=EUR|AFR
writeLines(c(
  paste(c("#Subpopulation order/codes: AFR=0", "EUR=1"), collapse = "\t"),
  paste(c("#chm","spos","epos","sgpos","egpos","n snps",
          "S1.0","S1.1","S2.0","S2.1","S3.0","S3.1"), collapse = "\t"),
  paste(c("19","1000","2000","0.0","0.1","2","0","1","1","0","0","1"), collapse = "\t"),
  paste(c("19","2000","3000","0.1","0.2","1","1","0","0","1","1","0"), collapse = "\t")
), msp_path)
```

```{r phased-pipeline}
result <- ancestry_split_phased(
  vcf_path  = vcf_path,
  msp_path  = msp_path,
  out_path  = out_path,
  write_vcf = FALSE,
  verbose   = FALSE
)

cat("African dosage (variants × samples):\n")
print(result$african)

cat("\nEuropean dosage:\n")
print(result$european)

cat("\nVariant info:\n")
print(result$variant_info)

# Sanity check: AFR + EUR == hap0 + hap1 everywhere
vcf_lines <- readLines(vcf_path)[-(1:3)]
raw <- do.call(rbind, lapply(vcf_lines, function(l) {
  gts <- strsplit(l, "\t")[[1]][10:12]
  vapply(gts, function(g) sum(as.integer(strsplit(g, "|", fixed=TRUE)[[1]])), integer(1))
}))
max_err <- max(abs((result$african + result$european) - raw))
cat("\nMax reconstruction error (must be 0):", max_err, "\n")
```

The output matrices have rows = variants and columns = samples.  With
`write_vcf = FALSE` no files are written; set it to `TRUE` (and provide a
real BCF path) to also produce `african_ds.gds` / `european_ds.gds` for the
downstream gene-level tests (see `vignette("gene-level-test")`).

The same call on a real cohort BCF + MSP (JHS chr19, 3313 samples, ~44 k
variants — runs when the raw data files are present):

```{r phased-pipeline-real, eval=file.exists("../../raw/LANTERN.phased.bcf")}
real_result <- ancestry_split_phased(
  vcf_path  = "../../raw/LANTERN.phased.bcf",
  msp_path  = "../../raw/LANTERN_chr19.msp.tsv.gz",
  out_path  = tempfile(),
  chrom     = "chr19",
  write_vcf = FALSE,
  verbose   = FALSE
)

cat("Variants kept:", nrow(real_result$african), "\n")
cat("Samples:      ", ncol(real_result$african), "\n")
cat("Monomorphic filtered:", real_result$overlap$n_monomorphic_filtered, "\n")
cat("First 3 variants:\n")
print(head(real_result$variant_info, 3))

# Sanity checks
cat("Min dosage (must be >= 0):", min(real_result$african, real_result$european), "\n")
cat("Max AFR dosage (must be <= 2):", max(real_result$african), "\n")
cat("Max EUR dosage (must be <= 2):", max(real_result$european), "\n")
```

---

## 3  Combined pipeline (`ancestry_split_combined`)

When you have a phased BCF and an RFMix MSP and want **both** phased and
unphased dosage estimates in one call, `ancestry_split_combined()` is the
single-function shortcut.  It parses the VCF and MSP once and returns a
named list with two elements, `$phased` and `$unphased`, each containing
`african`, `european`, `variant_info`, `sample_ids`, and `overlap`.

The synthetic VCF and MSP created above are still in scope.  The return
value is a named list: `$phased` and `$unphased`, each containing one
matrix per population (named by the MSP population labels) plus
`variant_info`, `sample_ids`, and `overlap`.

```{r combined-pipeline-synthetic}
combined <- ancestry_split_combined(
  vcf_path = vcf_path,
  msp_path = msp_path,
  verbose  = FALSE
)

# Population names come directly from the MSP header
pop_names <- setdiff(names(combined$phased),
                     c("variant_info", "sample_ids", "overlap"))
cat("Populations:", paste(pop_names, collapse = ", "), "\n\n")

cat("=== $phased ===\n")
for (p in pop_names) {
  cat(p, "dosage:\n"); print(combined$phased[[p]])
}

cat("\n=== $unphased ===\n")
for (p in pop_names) {
  cat(p, "dosage:\n"); print(combined$unphased[[p]])
}

cat("\nVariants in both modes:", nrow(combined$phased$variant_info), "\n")
cat("Samples:               ", length(combined$phased$sample_ids), "\n")
```

### 3.1  Three-population example

The same function works unchanged for K > 2.  Here we add a Native
American (NAT) population:

```{r combined-pipeline-3pop}
vcf3 <- file.path(tmp_dir, "phased3.vcf")
msp3 <- file.path(tmp_dir, "ancestry3.msp.tsv")

writeLines(c(
  "##fileformat=VCFv4.1",
  '##FORMAT=<ID=GT,Number=1,Type=String,Description="Genotype">',
  paste(c("#CHROM","POS","ID","REF","ALT","QUAL","FILTER","INFO",
          "FORMAT","S1","S2","S3"), collapse="\t"),
  paste(c("19","1500",".","A","G","100","PASS",".","GT","1|0","0|1","1|1"),
        collapse="\t"),
  paste(c("19","2500",".","C","T","100","PASS",".","GT","0|0","1|0","0|1"),
        collapse="\t")
), vcf3)

# Three populations: AFR=0, EUR=1, NAT=2
writeLines(c(
  paste(c("#Subpopulation order/codes: AFR=0", "EUR=1", "NAT=2"),
        collapse="\t"),
  paste(c("#chm","spos","epos","sgpos","egpos","n snps",
          "S1.0","S1.1","S2.0","S2.1","S3.0","S3.1"), collapse="\t"),
  # tract 1 [1000,2000): S1=AFR|EUR  S2=EUR|NAT  S3=AFR|NAT
  paste(c("19","1000","2000","0.0","0.1","1","0","1","1","2","0","2"),
        collapse="\t"),
  # tract 2 [2000,3000): S1=EUR|AFR  S2=AFR|EUR  S3=NAT|EUR
  paste(c("19","2000","3000","0.1","0.2","1","1","0","0","1","2","1"),
        collapse="\t")
), msp3)

combined3 <- ancestry_split_combined(vcf3, msp3, verbose = FALSE)

pop3 <- setdiff(names(combined3$phased),
                c("variant_info", "sample_ids", "overlap"))
cat("Populations:", paste(pop3, collapse = ", "), "\n\n")

cat("=== Phased dosages ===\n")
for (p in pop3) { cat(p, ":\n"); print(combined3$phased[[p]]) }

cat("\n=== Unphased dosages ===\n")
for (p in pop3) { cat(p, ":\n"); print(combined3$unphased[[p]]) }
```

The phased and unphased matrices share the same variant set here because
no extra variants become monomorphic under one mode but not the other.
For rare-variant-rich real cohorts the two `variant_info` frames may
differ by a handful of rows.

On the real JHS chr19 cohort (runs when raw files are present):

```{r combined-pipeline-real, eval=file.exists("../../raw/LANTERN.phased.bcf")}
combined_real <- ancestry_split_combined(
  vcf_path = "../../raw/LANTERN.phased.bcf",
  msp_path = "../../raw/LANTERN_chr19.msp.tsv.gz",
  chrom    = "chr19",
  verbose  = FALSE
)

pop_r <- setdiff(names(combined_real$phased),
                 c("variant_info", "sample_ids", "overlap"))

cat("=== Phased ===\n")
cat("Variants x samples:",
    nrow(combined_real$phased$variant_info), "x",
    length(combined_real$phased$sample_ids), "\n")
for (p in pop_r)
  cat(p, "total dosage:", round(sum(combined_real$phased[[p]]), 1), "\n")

cat("\n=== Unphased ===\n")
cat("Variants x samples:",
    nrow(combined_real$unphased$variant_info), "x",
    length(combined_real$unphased$sample_ids), "\n")
for (p in pop_r)
  cat(p, "total dosage:", round(sum(combined_real$unphased[[p]]), 1), "\n")

cat("\nNote: totals differ slightly because phased assigns alleles\n",
    "deterministically while unphased uses proportional splitting\n",
    "for ambiguous mixed heterozygotes.\n")
```

---

## 4  Choosing between unphased and phased

| | Unphased | Phased | Combined |
|---|---|---|---|
| **Ancestry input** | Diploid per-variant code (1/2/3) | Per-haplotype code from RFMix MSP | RFMix MSP (both derived automatically) |
| **Splitting rule** | Probabilistic (p1/p2 proportions) | Deterministic | Both |
| **Accuracy** | Good for intermediate-frequency variants | Better for rare variants | Side-by-side comparison |
| **Required tools** | None (pure R + C) | `bcftools` | `bcftools` |
| **Key function** | `ancestry_split_dosage()` | `ancestry_split_phased()` | `ancestry_split_combined()` |
